### 2025-10-13 YEAR2/AER210_LectureNotes/main.tex
oct 12

\section{Vector Calculus}
\section{Line Integrals}

@@

\chapter{Fluid Mechanics}
\end{document}

>>>>

\section{Vector Calculus}

@@

\begin{definition}[Surface Area]
    The surface area of a parametric surface \(\vec{r(u,v)}\) over a domain \(D\) is given by:
    \begin{equation}
        S = \iint_D ||\vec{r_u} \times \vec{r_v}|| \, dA
    \end{equation}
    where \(\vec{r_u} = \frac{\partial \vec{r}}{\partial u}\) and \(\vec{r_v} = \frac{\partial \vec{r}}{\partial v}\).
\end{definition}

\begin{example}[Surface Area of a Sphere]
    Find the surface area of a sphere with radius \(a\).

    We can parametrize the sphere using spherical coordinates:
    $$    
        \vec{r}(\phi, \theta) = \langle x, y , z \rangle = \langle a \sin \phi \cos \theta, a \sin \phi \sin \theta, a \cos \phi \rangle
    $$
    where over the domain \(D = \{(\phi, \theta) | 0 \leq \phi \leq \pi, 0 \leq \theta \leq 2\pi\}\).

    We can compute the partial derivatives:
    $$
        \vec{r_\phi} = \langle a \cos \phi \cos \theta, a \cos \phi \sin \theta, -a \sin \phi \rangle
    $$
    $$
        \vec{r_\theta} = \langle -a \sin \phi \sin \theta, a \sin \phi \cos \theta, 0 \rangle
    $$

    Then, we can compute the cross product:
    $$
        \vec{r_\phi} \times \vec{r_\theta} = 
        \begin{vmatrix}
            \hat{i} & \hat{j} & \hat{k} \\ 
            a\cos\phi\cos\theta & a\cos\phi\sin\theta & -a\sin\phi \\ 
            -a\sin\phi\sin\theta & a\sin\phi\cos\theta & 0
        \end{vmatrix}
        = a^2 \sin^2 \phi \cos \theta \hat{i} + a^2 \sin^2 \phi \sin \theta \hat{j} + a^2 \sin \phi \cos \phi \hat{k}
    $$

    The magnitude of the cross product is:
    $$
        ||\vec{r_\phi} \times \vec{r_\theta}|| = a^2 \sin \phi
    $$

    Finally, we can compute the surface area:
    $$
        S = \iint_D ||\vec{r_\phi} \times \vec{r_\theta}|| \, dA = \int_0^{2\pi} \int_0^\pi a^2 \sin \phi \, d\phi \, d\theta = 2\pi a^2 \left[ -\cos \phi \right]_0^\pi = 4\pi a^2
    $$

\subsection{Surface Integrals}
\begin{definition}[Surface Integral over Scalar Field]
    Consider a scalar field \(f(x,y,z)\) and a smooth parametric surface \(\vec{r(u,v)}\) over a domain \(D\). Then we define the surface integral of \(f\) over the surface as:
    \begin{equation}
        \iint_S f(x,y,z) \, dS = \iint_D f(\vec{r(u,v)}) ||\vec{r_u} \times \vec{r_v}|| \, du \, dv
    \end{equation}

    And if $S$ is piecewise smooth, then we can express the surface integral as the sum of the surface integrals over each smooth piece:
    \begin{equation}
        \iint_S f(x,y,z) \, dS = \sum_{i=1}^n \iint_{S_i} f(x,y,z) \, dS
    \end{equation}
    where \( \bigcup_{i=1}^n S_i = S \) and each \(S_i\) is a smooth surface.
\end{definition}
\begin{example}
    \textit{Evaluate the $\iint_S \sqrt{x^2 + y^2 + 1} \, dS$ where \(S\) is the surface given by $\vec{r}(u,v) = \langle u\cos v, u\sin v, v \rangle$, $0 \leq u \leq 1$, $0 \leq v \leq 2\pi$.}   
    We can compute the partial derivatives:
    $$        \vec{r_u} = \langle \cos v, \sin v, 0 \rangle $$
    $$        \vec{r_v} = \langle -u \sin v, u \cos v, 1 \rangle $$
    Then, we can compute the cross product:
    $$        \vec{r_u} \times \vec{r_v} = 
        \begin{vmatrix}
            \hat{i} & \hat{j} & \hat{k} \\ 
            \cos v & \sin v & 0 \\ 
            -u \sin v & u \cos v & 1
        \end{vmatrix}
        = \sin v \hat{i} - \cos v \hat{j} + u \hat{k}
    $$
    The magnitude of the cross product is:
    $$        ||\vec{r_u} \times \vec{r_v}|| = \sqrt{\sin^2 v + \cos^2 v + u^2} = \sqrt{1 + u^2} $$
    Finally, we can compute the surface integral:
    \begin{align*}
        \iint_S \sqrt{x^2 + y^2 + 1} \, dS &= \iint_D \sqrt{u^2 \cos^2 v + u^2 \sin^2 v + 1} \cdot \sqrt{1 + u^2} \, du \, dv \\
        &= \int_0^{2\pi} \int_0^1 \sqrt{u^2 + 1} \cdot \sqrt{1 + u^2} \, du \, dv = \int_0^{2\pi} \int_0^1 (1 + u^2) \, du \, dv \\
        &= \int_0^{2\pi} \left[ u + \frac{u^3}{3} \right]_0^1 \, dv = \int_0^{2\pi} \frac{4}{3} \, dv = \frac{8\pi}{3}
    \end{align*}
\end{example}
\begin{definition}[Orientability]
    A surface is said to be orientable if it is possible to assign a consistent normal vector at every point on the surface. In other words, a surface is orientable if it has two distinct sides. \\
    (Examples) Spheres, Torus, Cylinder, Plane \\    
    (Non-examples) Möbius Strip, Klein Bottle
\end{definition}

\begin{definition}[Orientation]
    An orientation of a surface is a choice of a continuous normal vector field on the surface. For a given surface, there are two possible orientations: one with the normal vector pointing "outward" and the other with the normal vector pointing "inward". The choice of orientation affects the sign of the surface integral. For closed surfaces, the outward orientation is typically chosen for obvious reasons.
\end{definition}
\begin{definition}[Surface Integrals of Vector Fields]
    Consider a vector field \(\vec{F}(x,y,z)\) and an orientable smooth surface \(S\) with a chosen orientation. Then we define the surface integral of \(\vec{F}\) over \(S\) as:
    \begin{equation}
        \iint_D \vec{F}(\vec{r(u,v)}) \cdot \hat{n} \, dS = \iint_D \vec{F}(\vec{r(u,v)}) \cdot (\vec{r_u} \times \vec{r_v}) \, du \, dv
    \end{equation}
    where $\hat{n} = \frac{\vec{r_u} \times \vec{r_v}}{||\vec{r_u} \times \vec{r_v}||}$ is the unit normal vector to the surface and $dS = ||\vec{r_u} \times \vec{r_v}|| \, du \, dv$ is the infinitesimal area element on the surface.
\end{definition}

\paragraph{Physical Interpretation - Flux Integration} Consider a fluid with density $\rho (x, y, z)$ and velocity filed $\vec{v(x, y, z)}$. For each patch of surface $dS$ with unit normal vector $\vec{n}$, the volume of fluid passing through the patch per unit time is given by is one of a prism given by:
$$
     \text{Height} = \vec{v} \cdot \vec{n} \, dt, \quad \text{Volume} = (\vec{v} \cdot \vec{n}) \, dS \, dt \, \implies dm = \rho (\vec{v} \cdot \vec{n}) \, dS \, dt
$$
The mass flow rate $d\text{Mass}/dt$ is given by:
$$
    \frac{\text{Mass}}{dt} = \rho (\vec{v} \cdot \vec{n}) \, dS
$$
So we write:
$$
    \text{Mass flow rate} = \iint_S \rho \vec{v} \cdot \vec{n}\, dS = \iint_S F \cdot \vec{n} \, dS = \iint_S \vec{F} \cdot d\vec{S}
$$
where $\vec{F} = \rho \vec{v}$ is the flux density vector field.

\begin{example}
    \textit{Calculate the flux of the vector filed $\vec{F}(x,y,z) = x \, \hat{i} + y \, \hat{j} + z \, \hat{k}$ across the closed cylindrical surface \(S\) given by \(x^2 + y^2 = a^2\), \(-1 \leq z \leq 1\), oriented outward.}

    We first parametrize the cylindrical surface as:
    $$        \vec{r}(\theta, z) = \langle a \cos \theta, a \sin \theta, z \rangle $$
    over the domain \(D = \{(\theta, z) | 0 \leq \theta \leq 2\pi, -1 \leq z \leq 1\}\).

    Then, we can compute the partial derivatives:
    $$        \vec{r_\theta} = \langle -a \sin \theta, a \cos \theta, 0 \rangle $$
    $$        \vec{r_z} = \langle 0, 0, 1 \rangle $$
    Then, we can compute the cross product:
    $$        \vec{r_\theta} \times \vec{r_z} = 
        \begin{vmatrix}
            \hat{i} & \hat{j} & \hat{k} \\ 
            -a \sin \theta & a \cos \theta & 0 \\ 
            0 & 0 & 1
        \end{vmatrix}
        = a \cos \theta \hat{i} + a \sin \theta \hat{j} + 0 \hat{k}
    $$
    The magnitude of the cross product is:
    $$        ||\vec{r_\theta} \times \vec{r_z}|| = \sqrt{(a \cos \theta)^2 + (a \sin \theta)^2 + 0^2} = a $$
    Finally, we can compute the surface integral:
    \begin{align*}
        \iint_S \vec{F} \cdot d\vec{S} &= \iint_D \vec{F}(\vec{r}(\theta, z)) \cdot (\vec{r_\theta} \times \vec{r_z}) \, d\theta \, dz \\
        &= \int_0^{2\pi} \int_{-1}^1 \langle a \cos \theta, a \sin \theta, z \rangle \cdot \langle a \cos \theta, a \sin \theta, 0 \rangle \, dz \, d\theta \\
        &= \int_0^{2\pi} \int_{-1}^1 a^2 (\cos^2 \theta + \sin^2 \theta) \, dz \, d\theta = 2\int_0^{2\pi} a^2 (1) (1 - 0) \, d\theta = 4\pi a^2
    \end{align*}
    And for the top and bottom surfaces, we have:
    \begin{align*}
        \iint_{S_{top}} \vec{F} \cdot d\vec{S} &= \iint_D \vec{F}(\vec{r}(r, \theta)) \cdot (\vec{r_r} \times \vec{r_\theta}) \, dr \, d\theta \\
        &= \int_0^{2\pi} \int_0^a \langle r \cos \theta, r \sin \theta, 1 \rangle \cdot \langle 0, 0, r \rangle \, dr \, d\theta \\
        &= \int_0^{2\pi} \int_0^a r \, dr \, d\theta = \int_0^{2\pi} \left[ \frac{r^2}{2} \right]_0^a \, d\theta = \int_0^{2\pi} \frac{a^2}{2} \, d\theta = \pi a^2
    \end{align*}
    And then, by symmetry, we have:
    $$        \iint_{S_{bottom}} \vec{F} \cdot d\vec{S} = \pi a^2 $$
    So, the total flux across the closed surface is:
    $$        \iint_S \vec{F} \cdot d\vec{S} = 4\pi a^2 + \pi a^2 + \pi a^2 = 6\pi a^2 $$
\end{example}

\begin{example}
    Find the flux of $\vec{F}(x,y,z) = \frac{2x \hat{i} + 2y \hat{j}}{x^2 + y^2} + \hat{k}$ across the surface of the paraboloid $\vec{r}(u, v) = \langle u \cos v, u \sin v, u^2 \rangle$, $0 \leq u \leq 1$, $0 \leq v \leq 2\pi$, oriented downward.

    We can compute the partial derivatives:
    $$        \vec{r_u} = \langle \cos v, \sin v, 2u \rangle $$
    $$        \vec{r_v} = \langle -u \sin v, u \cos v, 0 \rangle $$
    Then, we can compute the cross product:
    $$        \vec{r_u} \times \vec{r_v} = -2u^2 \cos v \hat{i} - 2u^2 \sin v \hat{j} + u \hat{k}
    $$
    Notice, since the surface is oriented downward, we take the negative of the cross product:
    $$        \vec{r_v} \times \vec{r_u} = 2u^2 \cos v \hat{i} + 2u^2 \sin v \hat{j} - u \hat{k}
    $$
\end{example}
\chapter{Fluid Mechanics}
\section{Introduction to Fluid Mechanics}
\begin{definition}[Fluid]
    A fluid is a substance that continuously deforms (flows) under an applied shear stress, no matter how small the stress. Fluids include liquids and gases.

\end{definition}
\end{document}

### 2025-10-13 YEAR2/ECE253_LectureNotes/main.tex
oct 12

>>>>

\begin{example}[Alternative Design]
    Alternatively, we can design the state diagram in which each state represents the last three values. We can set the following states table:
    \begin{table}[h!]
        \centering
        \begin{tabular}{|c|cc|c|}
            \hline
            $y_1 y_2 y_3$ & $w=0$ & $w=1$ & Output \\
            \hline
            000 & 000 & 001 & 0 \\
            001 & 010 & 011 & 0 \\
            010 & 100 & 101 & 0 \\
            011 & 110 & 111 & 0 \\
            100 & 000 & 001 & 0 \\
            101 & 010 & 011 & 1 \\
            110 & 100 & 101 & 0 \\
            111 & 110 & 111 & 0 \\
            \hline
        \end{tabular}
        \caption{State Table for Alternative FSM that detects `101'}
        \label{tab:alternative_fsm_state_table}
    \end{table}
    And then we can draw the following equations:
    \begin{itemize}
        \item $Y_3 = w$
        \item $Y_2 = y_3$
        \item $Y_1 = y_2$
    \end{itemize}
    The output $Z = y_3 \overline{y_2} y_1$.
    This design is simpler, but it uses more states (8 states instead of 4 states). This is an example of the trade-off between the number of states and the complexity of the logic.

    We call this a shift register, where the input is shifted into the register on each clock cycle. The output is taken from the last bit of the register.
\end{example}

### 2025-10-13 YEAR2/MAT292_LectureNotes/main.tex
oct 12

Therefore the general solution of \(\mathbf{x}'=A\mathbf{x}\) is the linear combination
\begin{equation}
\mathbf{x}(t)=C_1\,x_1(t) + C_2\,x_2(t)
= e^{\lambda t} \begin{bmatrix}
C_1 + C_2 t \\[4pt] C_2
\end{bmatrix}
\end{equation}
The phase portrait of this system is shown below:
% insert later

\begin{definition}[Trace-Determinant Plane]
    We found that we can find the eigenvalues of a \( 2 \times 2 \) matrix \( A \) using:
    $$
    \lambda_{1,2} = \frac{\text{tr}(A) \pm \sqrt{\text{tr}(A)^2 - 4\det(A)}}{2}
    $$
    where
    \begin{itemize}
        \item \( \text{tr}(A) = \lambda_1 + \lambda_2 = a_{11} + a_{22} \) is the trace of \( A \)
        \item \( \det(A) = \lambda_1 \lambda_2 = a_{11}a_{22} - a_{12}a_{21} \) is the determinant of \( A \)   
    \end{itemize}
    And we have the following analysis:
    \begin{itemize}
        \item If \( \det(A) < 0 \) and \( \text{tr}(A) > 0 \), then \( \lambda_1 > 0 > \lambda_2 \) (saddle node).
        \item Consider the discriminant boundry \( \text{tr}(A)^2 - 4\det(A) = 0 \), it is a parabola that upward opening with vertex at the origin. Consider region above that, where there is complex eigenvalues:
        \begin{itemize}
            \item If \( \text{tr}(A) > 0 \) and \( \det(A) > 0 \), then \( \alpha = \text{tr}(A)/2 > 0 \) (spiral source).
            \item If \( \text{tr}(A) < 0 \) and \( \det(A) > 0 \), then \( \alpha = \text{tr}(A)/2 < 0 \) (spiral sink).
            \item If \( \text{tr}(A) = 0 \) and \( \det(A) > 0 \), then \( \lambda_1 = -\lambda_2 = i\beta \) (center).
        \end{itemize}
        \item Consider the region below the discriminant boundry, where there is real eigenvalues:
        \begin{itemize}
            \item If \( \text{tr}(A) > 0 \) and \( \det(A) > 0 \), then \( \lambda_1 \geq \lambda_2 > 0 \) (nodal source).
            \item If \( \text{tr}(A) < 0 \) and \( \det(A) > 0 \), then \( 0 > \lambda_1 \geq \lambda_2 \) (nodal sink).
        \end{itemize}
        \item \textbf{Degenerate Star Node} Consider the line \( \text{tr}(A)^2 - 4\det(A) = 0 \).
        \item \textbf{Lines of Nodes} Consider the line \( \det(A) = 0 \).
    \end{itemize}
\end{definition}

@@

\begin{theorem}[Pomcor\'e-Bendixson Theorem]
    Consider a two-dimensional autonomous system:
    $$\begin{cases}
        x' = f(x, y) \\
        y' = g(x, y)
    \end{cases}$$
    where \( f \) and \( g \) are continuously differentiable functions. If there exists a trajectory that remains in a closed, bounded region \( R \) of the phase plane for all future time and if \( R \) contains only unstable nodes, then all trajectories in \( R \) approach a limit cycle as \( t \to \infty \).
\end{theorem}
\paragraph{Intuition} A monotonically increasing function that gets bounded must converge.

\section{Matrices and Linear Algebra}

\subsection{Matrix Exponentials}
\begin{shaded}
\subsubsection*{Taylor Series Definition of Exponentials}
Consider a scalar exponential, then,
$$e^x = \sum_{n=0}^{\infty} \frac{x^n}{n!}$$
is the Taylor series expansion of \( e^x \) around \( x = 0 \). This series converges for all real numbers \( x \).
\end{shaded}

@@

>>>>

Therefore the general solution of \(\mathbf{x}'=A\mathbf{x}\) is the linear combination
\begin{equation}
\mathbf{x}(t)=C_1\,x_1(t) + C_2\,x_2(t)
= e^{\lambda t} \begin{bmatrix}
C_1 + C_2 t \\[4pt] C_2
\end{bmatrix}
\end{equation}

\paragraph{Applying the matrix to the generalized eigenvector} We can apply the matrix \( A \) to the generalized eigenvector \( v_g \), then , $Av_g$ would stay in the span of $v$ and $v_g$: In other words, we call this subspace invariant under the transformation $A$.

\begin{example}[Generalized Eigenvectors]
    Consider the system of ODEs:
    $$
    \textbf{x}' = \begin{bmatrix} 4 & 1 \\ -1 & 2
    \end{bmatrix}\textbf{x}
    $$
    We find the eigenvalues and eigenvectors by solving the characteristic polynomial:
    \begin{align*}
    \det(A - \lambda I) &= \det\begin{bmatrix} 4 - \lambda & 1 \\ -1 & 2 - \lambda
    \end{bmatrix} \\
    &= (\lambda - 3)^2 = 0 \\
    \implies \lambda_1 &= 3, \quad \lambda_2 = 3
    \end{align*}
    The corresponding eigenvector is:
    \begin{align*}
    \vec{v}_1 &= \begin{bmatrix} 1 \\ -1
    \end{bmatrix}
    \end{align*}
    Since there is only one linearly independent eigenvector, we need to find a generalized eigenvector \( v_g \) satisfying:
    $$
    (A - 3I)v_g = v_1
    $$
    Solving this, we find:
    $$
    v_g = \begin{bmatrix} 0 \\ 1
    \end{bmatrix}
    $$
    Thus, the two linearly independent solutions to the system are:
    $$x_1(t) = v_1 e^{3t} = \begin{bmatrix} 1 \\ -1
    \end{bmatrix} e^{3t}$$
    $$x_2(t) = (v_g + tv_1)e^{3t} = \left(\begin{bmatrix} 0 \\ 1
    \end{bmatrix} + t\begin{bmatrix} 1 \\ -1
    \end{bmatrix}\right)e^{3t}$$

    Therefore, the general solution to the system is:
    $$\textbf{x}(t) = c_1 x_1(t) + c_2 x_2(t)$$
    where \( c_1 \) and \( c_2 \) are constants determined by the initial conditions.
\end{example}
\begin{definition}[Trace-Determinant Plane]
    We found that we can find the eigenvalues of a \( 2 \times 2 \) matrix \( A \) using:
    $$
    \lambda_{1,2} = \frac{\text{tr}(A) \pm \sqrt{\text{tr}(A)^2 - 4\det(A)}}{2}
    $$
    where
    \begin{itemize}
        \item \( \text{tr}(A) = \lambda_1 + \lambda_2 = a_{11} + a_{22} \) is the trace of \( A \)
        \item \( \det(A) = \lambda_1 \lambda_2 = a_{11}a_{22} - a_{12}a_{21} \) is the determinant of \( A \)   
    \end{itemize}
    And we have the following analysis:
    \begin{itemize}
        \item If \( \det(A) < 0 \) and \( \text{tr}(A) > 0 \), then \( \lambda_1 > 0 > \lambda_2 \) (saddle node).
        \item Consider the discriminant boundry \( \text{tr}(A)^2 - 4\det(A) = 0 \), it is a parabola that upward opening with vertex at the origin. Consider region above that, where there is complex eigenvalues:
        \begin{itemize}
            \item If \( \text{tr}(A) > 0 \) and \( \det(A) > 0 \), then \( \alpha = \text{tr}(A)/2 > 0 \) (spiral source).
            \item If \( \text{tr}(A) < 0 \) and \( \det(A) > 0 \), then \( \alpha = \text{tr}(A)/2 < 0 \) (spiral sink).
            \item If \( \text{tr}(A) = 0 \) and \( \det(A) > 0 \), then \( \lambda_1 = -\lambda_2 = i\beta \) (center).
        \end{itemize}
        \item Consider the region below the discriminant boundry, where there is real eigenvalues:
        \begin{itemize}
            \item If \( \text{tr}(A) > 0 \) and \( \det(A) > 0 \), then \( \lambda_1 \geq \lambda_2 > 0 \) (nodal source).
            \item If \( \text{tr}(A) < 0 \) and \( \det(A) > 0 \), then \( 0 > \lambda_1 \geq \lambda_2 \) (nodal sink).
        \end{itemize}
        \item \textbf{Degenerate Star Node} Consider the line \( \text{tr}(A)^2 - 4\det(A) = 0 \).
        \item \textbf{Lines of Nodes} Consider the line \( \det(A) = 0 \).
    \end{itemize}
\end{definition}

@@

\begin{theorem}[Pomcor\'e-Bendixson Theorem]
    Consider a two-dimensional autonomous system:
    $$\begin{cases}
        x' = f(x, y) \\
        y' = g(x, y)
    \end{cases}$$
    where \( f \) and \( g \) are continuously differentiable functions. If there exists a trajectory that remains in a closed, bounded region \( R \) of the phase plane for all future time and if \( R \) contains only unstable nodes, then all trajectories in \( R \) approach a limit cycle as \( t \to \infty \).
\end{theorem}
\paragraph{Intuition} A monotonically increasing function that gets bounded must converge. A similar idea applies to trajectories in a dynamical system: if they are confined to a bounded region and cannot escape, they must eventually settle into a steady state, such as a limit cycle.

\section{Matrix Exponentials}
\begin{shaded}
\subsubsection*{Taylor Series Definition of Exponentials}
Consider a scalar exponential, then,
$$e^x = \sum_{n=0}^{\infty} \frac{x^n}{n!}$$
is the Taylor series expansion of \( e^x \) around \( x = 0 \). This series converges for all real numbers \( x \).
\end{shaded}

@@

\paragraph{What if \( A \) is not diagonalizable?} If \( A \) is not diagonalizable there could be two cases:
\begin{itemize}
    \item If we know the fundamental set of solutions \( \{x_1(t), x_2(t), \ldots, x_n(t)\} \), then we can construct the fundamental matrix \( \Phi(t) \) and compute the matrix exponential as:
    \begin{equation}
    e^{At} = \Phi(t)\Phi(0)^{-1}
    \end{equation}
    where \( \Phi(0) \) is the fundamental matrix evaluated at \( t = 0 \). And $\phi = [x_1, x_2, \ldots, x_n]$.
    \item Otherwise, we can consider its Jordan Canonical Form.
\end{itemize}

\paragraph{Jordan Canonical Form} We first define the AM and GM of an eigenvalue:
\begin{definition}[Geometric and Algebraic Multiplicity]
    Consider a square matrix \( A \in \mathbb{R}^{n \times n} \) and an eigenvalue \( \lambda \) of \( A \). The algebraic multiplicity (AM) of \( \lambda \) is the number of times \( \lambda \) appears as a root of the characteristic polynomial of \( A \). The geometric multiplicity (GM) of \( \lambda \) is the dimension of the eigenspace corresponding to \( \lambda \), which is the null space of \( A - \lambda I \). Diagonalizability is characterized by the condition that for each eigenvalue, its algebraic multiplicity equals its geometric multiplicity.
\end{definition}

\begin{definition}[Jordan Block]
    A Jordan block \( J_k(\lambda) \) of size \( k \) associated with an eigenvalue \( \lambda \) is a \( k \times k \) matrix of the form:
    \begin{equation}
    J_k(\lambda) = \begin{bmatrix}
    \lambda & 1 & 0 & \cdots & 0 \\
    0 & \lambda & 1 & \cdots & 0 \\
    0 & 0 & \lambda & \cdots & 0 \\
    \vdots & \vdots & \vdots & \ddots & 1 \\
    0 & 0 & 0 & \cdots & \lambda
    \end{bmatrix}
    \end{equation}
    where \( \lambda \) is the eigenvalue and the superdiagonal entries are all ones, while all other entries are zero.

\begin{definition}[Jordan Canonical Form]
    A square matrix \( A \in \mathbb{R}^{n \times n} \) is said to be in Jordan canonical form if it is a block diagonal matrix where each block is a Jordan block. Specifically, \( A \) can be expressed as:
    \begin{equation}
    A = PJP^{-1}
    \end{equation}
    where \( P \) is an invertible matrix and \( J \) is the Jordan matrix composed of Jordan blocks along its diagonal. The Jordan canonical form provides a way to represent matrices that are not diagonalizable, allowing for the computation of functions of matrices, such as the matrix exponential.
\end{definition}
\begin{example}
    Consider the system of ODEs:
    $$x' = \begin{bmatrix}
    1 & -1 \\
    1 & 3
    \end{bmatrix} x$$
    The characteristic polynomial is given by:
    \begin{align*}
        p(\lambda) &= \det\begin{bmatrix}
        1 - \lambda & -1 \\
        1 & 3 - \lambda
        \end{bmatrix} \\
        &= (1 - \lambda)(3 - \lambda) + 1 \\
        &= \lambda^2 - 4\lambda + 4 \\
        &= (\lambda - 2)^2
    \end{align*}
    So, we have a repeated eigenvalue \( \lambda = 2 \) with algebraic multiplicity 2. To find the eigenvectors, we solve:
    \begin{align*}
        (A - 2I)v &= 0 \\
        \begin{bmatrix}
        -1 & -1 \\
        1 & 1
        \end{bmatrix} \begin{bmatrix}
        v_1 \\ v_2
        \end{bmatrix} &= 0
    \end{align*}
    This gives us the eigenvector \( v = \begin{bmatrix}1 \\ -1 \end{bmatrix} \). Since the geometric multiplicity is 1, we need to find a generalized eigenvector \( w \) such that:
    \begin{align*}
        (A - 2I)w &= v \\
        \begin{bmatrix}
        -1 & -1 \\
        1 & 1
        \end{bmatrix} \begin{bmatrix}
        w_1 \\ w_2
        \end{bmatrix} &= \begin{bmatrix}1 \\ -1
        \end{bmatrix}
    \end{align*}
    Solving this, we can take \( w = \begin{bmatrix}0 \\ -1 \end{bmatrix} \). Now, we can form the Jordan matrix:
    $$
    J = \begin{bmatrix}
    2 & 1 \\
    0 & 2
    \end{bmatrix} = 2I + \begin{bmatrix}
    0 & 1 \\
    0 & 0
    \end{bmatrix}
    $$
    Obserbe that $N = \begin{bmatrix}
    0 & 1 \\
    0 & 0
    \end{bmatrix}$ is nilpotent since $N^2 = 0$. And that $I$ commutes with everything. So, we can compute the matrix exponential as:
    \begin{align*}
        e^{At} &= Pe^{Jt}P^{-1} \\
        &= P e^{(2I + N)t} P^{-1} \\
        &= P e^{2tI} e^{Nt} P^{-1} \\
        &= P e^{2t} \left(I + Nt + \frac{(Nt)^2}{2!} + \ldots \right) P^{-1} \\
        &= P e^{2t} \left(I + Nt\right) P^{-1} \\
        &= e^{2t} P \begin{bmatrix}
        1 & t \\
        0 & 1
        \end{bmatrix} P^{-1}
    \end{align*}
    where \( P = \begin{bmatrix}1 & 0 \\
    -1 & -1
    \end{bmatrix} \) and \( P^{-1} = \begin{bmatrix}
    1 & 0 \\
    -1 & -1
    \end{bmatrix} \). Thus, we have:
    $$
        e^{At} = e^{2t} \begin{bmatrix}1 + t & t \\
        -t & 1 - t
        \end{bmatrix}
    $$
\end{example}

\section{Higher Order ODEs}
\subsection{Second Order Linear ODEs}

\begin{definition}[General Form of Second Order Linear ODEs]
    A second-order linear ordinary differential equation (ODE) has the general form:
    \begin{equation}
     y'' = f(t, y, y')
    \end{equation}

\begin{example} \label{ex:2ndOrderTo1stOrder}
    We can convert a second-order linear ODE into a system of first-order ODEs. Consider the second-order ODE:
    $$
     y'' + p(t)y' + q(t)y = g(t)
    $$
    We can define:
    $$\begin{cases}
    x_1 = y \\
    x_2 = y'
    \end{cases}$$
    Then, we have the system of first-order ODEs:
    $$\begin{cases}
    x_1' = x_2 \\
    x_2' = -g(t) - p(t)x_2 - q(t)x_1
    \end{cases}$$

    In matrix form, this can be written as:
    $$\begin{bmatrix}
    x_1' \\ x_2'
    \end{bmatrix} = \begin{bmatrix}
    0 & 1 \\
    -q(t) & -p(t)
    \end{bmatrix} \begin{bmatrix}
    x_1 \\ x_2
    \end{bmatrix} + \begin{bmatrix}
    0 \\ -g(t)
    \end{bmatrix}$$ 
    This transformation allows us to apply techniques for first-order systems to analyze and solve second-order ODEs.

\begin{example}
    Consider the above Example \ref{ex:2ndOrderTo1stOrder} with constant coefficients:
    $$y'' + p y' + q y = 0
    $$
    We can write the corresponding system of first-order ODEs as:
    $$\begin{bmatrix}
    x_1' \\ x_2'
    \end{bmatrix} = \begin{bmatrix}
    0 & 1 \\
    -q & -p
    \end{bmatrix} \begin{bmatrix}
    x_1 \\ x_2
    \end{bmatrix}
    $$
    And you should be able to solve and analyze the system using the techniques we have learned for first-order systems, especially the trace-determinant plane.

\begin{definition}[Characteristic Equation]
    The characteristic equation of a second-order linear ODE with constant coefficients:
    $$y'' + p y' + q y = 0$$
    is given by the quadratic equation:
    \begin{equation}
    r^2 + pr + q = 0
    \end{equation}
    where \( r \) represents the roots of the characteristic polynomial. The nature of the roots (real and distinct, real and repeated, or complex conjugates) determines the form of the general solution to the ODE. This comes from the assuned solution of the form \( y = e^{rt} \).
\end{definition}

\begin{example}[Cauchy-Euler Equation]
    Consider the following second-order ODE:
    $$t^2 y'' - t y' + y = 0$$
    This is in the form of a Cauchy-Euler equation:
    $$
    \sum_m y^{(m)}(t) a_m t^m = 0
    $$
    where \( a_2 = 1, a_1 = -1, a_0 = 1 \). We can solve this by assuming a solution of the form \( y = t^r \). Then, we have:
    \begin{align*}
        y' &= r t^{r-1} \\
        y'' &= r(r-1) t^{r-2}
    \end{align*}
    Substituting these into the original equation gives:
    \begin{align*}
        t^2 (r(r-1) t^{r-2}) - t (r t^{r-1}) + t^r &= 0 \\
        r(r-1) t^r - r t^r + t^r &= 0 \\
        (r^2 - 2r + 1) t^r &= 0 \\
        (r - 1)^2 t^r &= 0
    \end{align*}
    Thus, we have a repeated root \( r = 1 \). The general solution for this case is given by:
    $$y(t) = C_1 t + C_2 t \ln(t)$$
    where \( C_1 \) and \( C_2 \) are constants determined by initial conditions.
\end{example}
